Handling String-Number Conversion in String Constraint Solving

### 2019-11-18 draft.tex
wording

Symbolic execution is a very popular technique that allows the programmers to check the feasibility of a path in a given program (i.e., determining the value of the inputs under which the given path can be executed) \cite{????}).
The path feasibility problem is usually solved by a reduction to the satisfiability problem of a formula. More precisely, program statements in the path are usually translated to equivalent constraints in static single assignment (SSA) form and then solved by \emph{satisfiability module theory (SMT)}  solvers. The types of constraints that needed to be solved by the SMT solvers depend on the type of program variables. Therefore, there is a need for SMT solvers that support different combinations of theories (i.e., depending on the types of the manipulated variables).

Among all types of program variables, \emph{string data type}  is omnipresent in modern programming languages.  Various infamous security vulnerabilities such as injection and cross-site scripting attack are caused by malicious string values.  Therefore, string constraint solving has received considerable attention in the constraint solving community.
Operations such as \emph{equality constraints} (e.g. $x.y = y.x$), \emph{regular constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$), are widely supported by most state-of-the-art string constraint solvers such as Z3  \cite{?}, CVC4  \cite{?}, Z3Str3  \cite{?}, and Trau+ \cite{?}. 
 However,   solvers have limited support for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often very limited. Thus, $n = \sti{\\texttt{"2"}}$  or $x = \its{5}$  are handled but $n = \sti{x}$ (i.e., converting the string $x$ to the integer $n$) and  $x = \its{n}$ (i.e.,  converting the integer $n$ to the string $x$)
are ones of the core operations that require stronger solver support.

@@

In fact, for many languages such as Java, the string-number conversion operations are is a niche operation, and much symbolic execution can not  be done without supporting it. For example, the program fragment below is a variant of the Luhn test algorithm \cite{?} that is often used in credit card or personal ID validation.

@@

Here $value_0$ and $last_0$ are string variables and the others are integer variables. The method $\chatAt{x}{i}$ returns the character at the index $i$ in the string $x$ while $n=\ite{b}{e}{e'}$ assigns to $n$ the value of the expression $e$ if $b$ is true and the value of the expression $e'$ otherwise. Line 1 describes the initial condition: \textsf{value} should be a sequence of digits and \textsf{sum} is initially zero. Lines 2-4 and lines 5-7 describe one execution of the first and second loop, respectively. Line 8 describes the condition of $i_2$ before leaving the loop. Finally, Lines 9-10 describe the condition that the last digit of sum is zero. Observe that to describe such a program path, we need a solver that supports the following types of constraints:
\begin{itemize}
	\item \emph{regular} constraints (e.g., $value_0 \in [1,9]^+$, which says $value_0$ is in the regular language $[0,9]^+$),
	\item \emph{integer} constraints (e.g., $i_0 = |value_0| -1$, which says $i_0$ equals the length of $value_0$ minus one),
	\item \emph{equality} constraints (often $y=\chatAt{x}{i}$ is encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$, which uses equality of string terms $x$ and $x_1.x_2.x_3$), and
	\item \emph{string-number conversion} (e.g., $\its{sum_2}$, which is the string value of the number $sum_2$).
\end{itemize}

Currently, even for a simple constraint like the above one, most solvers already fail to provide a correct answer. Therefore, there is an urgent need for developing efficient techniques for solving the combination of above constraints. To make matters worse, many scripting languages, of which JavaScript is the most notable ---it powers most interactive content on the Web, and increasingly server-side code with Node.js---, has conversions between strings and integers embedded in the core of its semantics. To see how, consider the following program:

@@

For proving UNSAT, our framework does not set any restriction on which decision procedure to use. One can select any decision procedure they want to use and ``over-approximate'' the input constraint to one that falls in the decidable fragment, e.g., the acyclic fragment~\cite{abdulla2014string}, the straight-line fragment~\cite{chen2019decision}, or the chain-free fragment~\cite{abdulla2019chain}. For instance, we can convert $xy=yx$ to two formulas $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1  \Leftrightarrow x\not\in [0-9]^*$, where $\Leftrightarrow$ means ``if and only if''. This produce new formulas that over-approximate the original ones and fall in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Since $\Phi$ is satisfiable, it is expected that the over-approximation module will not return UNSAT and we proceed to the the under-approximation module to prove SAT.

In our implementation \tool, we further add heuristics trying to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some other cases, we can derive $n=12$ from some integer constraints. In such case, we can further derive that the value of $x$ is in the regular language $(0^*12)$.

For proving SAT, our procedure first restrict the domain of each string variables to the language defined by some parametric flat automata (PFA) and then convert the string constraints under the domain restriction to equisatisfiable linear formulas.  Here we try to explain what the linear formula looks like by an example. Assume that we project the domain of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_1$, $v_2$, $v_3$, $v_4$ in the figure are \emph{character} variables. One special treatment we do in this paper is that we encoding characters as natural numbers. So $v_1$, $v_2$, $v_3$, $v_4$ are also integer variables. For example, from the formula $ y \in (12)^+$, we may derive $v_3=1$ and $v_4=2$.

@@

\yfc{maybe add a flowchart later if space allows}

>>>>

Symbolic execution is a very popular technique that allows programmers to check the feasibility of a path in a given program, i.e., determining the value of the inputs under which the given path can be executed. \cite{????}).
The path feasibility problem is usually solved by a reduction to the satisfiability of a formula. More precisely, program statements in the path are usually translated to equivalent constraints in static single assignment (SSA) form and then solved by \emph{satisfiability module theory (SMT)}  solvers. The types of constraints needed depend on the types of program expressions to be analyzed; therefore, SMT solvers must support different combinations of theories to handle a wide range of types.

In particular, \emph{string data}  is omnipresent in modern programming languages, especially in Web code that gets input as strings.  Various infamous security vulnerabilities, such as injection and cross-site scripting attack, can be caused by malicious string values.  Therefore, solving string constraint has received considerable attention in the constraint solving community.
Operations such as \emph{equality constraints} (e.g. $x.y = y.x$), \emph{regular constraints} (e.g., $x \in (a.b)^*$), and \emph{integer constraints} (e.g., $|x|-|y|>3$), are widely supported by most state-of-the-art string constraint solvers such as Z3  \cite{?}, CVC4  \cite{?}, Z3Str3  \cite{?}, and Trau+ \cite{?}. 
 However,   solvers have limited support for operations that convert between strings and integers; string length operations are sometimes well supported, but not parsing a string as an integer or turning an integer value into its string form.  When there is support, it is often very limited. Thus, $n = \sti{\\texttt{"2"}}$  or $x = \its{5}$  are handled but $n = \sti{x}$ (i.e., converting the string $x$ to the integer $n$) and  $x = \its{n}$ (i.e.,  converting the integer $n$ to the string $x$)
are core operations that require stronger solver support to handle real code.

@@

In fact, for many languages such as Java, string-number conversions are a niche operation, and much symbolic execution has been done without supporting it.  However, code that receives input tends to need to convert at least some of that input into numbers, which makes string-integer conversions critical to checking it.  For example, the program fragment below is a variant of the Luhn test algorithm \cite{?} that is often used in credit card or personal ID validation.

@@

Here $value_0$ and $last_0$ are string variables and the others are integer variables. The method $\chatAt{x}{i}$ returns the character at index $i$ in the string $x$ while $n=\ite{b}{e}{e'}$ assigns to $n$ the value of the expression $e$ if $b$ is true and the value of the expression $e'$ otherwise. Line 1 describes the initial condition: \textsf{value} should be a sequence of digits and \textsf{sum} is initially zero. Lines 2-4 and lines 5-7 describe one execution of the first and second loop, respectively. Line 8 describes the condition of $i_2$ before leaving the loop. Finally, Lines 9-10 describe the condition that the last digit of $sum$ is zero. Observe that to describe such a program path, we need a solver that supports the following types of constraints:
\begin{itemize}
	\item \emph{regular} constraints (e.g., $value_0 \in [1,9]^+$, which says $value_0$ is in the regular language $[0,9]^+$),
	\item \emph{integer} constraints (e.g., $i_0 = |value_0| -1$, which says $i_0$ equals the length of $value_0$ minus one),
	\item \emph{equality} constraints (often $y=\chatAt{x}{i}$ is encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$, which uses equality of string terms $x$ and $x_1.x_2.x_3$), and
	\item \emph{string-number conversion} (e.g., $\its{sum_2}$, which is the string value of the number $sum_2$).
\end{itemize}

Currently, even for the above constraint, which is from a simple program, most solvers already fail to provide a correct answer. Therefore, there is an urgent need for developing efficient techniques for solving such combinations of constraints. To make matters worse, many scripting languages, of which JavaScript is the most notable ---it powers most interactive content on the Web, and increasingly server-side code with Node.js---, has conversions between strings and integers embedded in the core of its semantics. To see how, consider the following program:

@@

For proving UNSAT, our framework does not set any restriction on which decision procedure to use. One can select any decision procedure they want to use and ``over-approximate'' the input constraint to one that falls in the decidable fragment, e.g., the acyclic fragment~\cite{abdulla2014string}, the straight-line fragment~\cite{chen2019decision}, or the chain-free fragment~\cite{abdulla2019chain}. For instance, we can convert $xy=yx$ to two formulas $\{x_1=xy, x_2=yx\}$ using two auxiliary variables $x_1,x_2$ and replace the constraint $n=\sti{x}$ with $n=-1 \vee (n\neq -1 \wedge x\in [0-9]^*$). This produce new formulas that over-approximate the original ones and fall in the chain-free fragment~\cite{abdulla2019chain}, which is decidable. Since $\Phi$ is satisfiable, the over-approximation module will not return UNSAT. In our implementation \tool, we further add heuristics trying to derive the constant value of any side of the constraint $n=\sti{x}$ to refine the over-approximation. For instance, assume in some other cases, we can derive the value of $n=12$ from some integer constraints. In such case, we can further derive the value of $x$ is in the regular language $(0^*12)$.

Then our procedure will restrict the domain of each string variables to the language defined by some parametric flat automata (PFA) and convert them to equisatisfiable linear formula. The details of the under-approximation component will be described in Section~\ref{section:under_approximate}. Here we try to explain what the linear formula looks like by an example. Assume that we project the domain of $x$ and $y$ to the PFA in Figure~\ref{fig:sfa} (a) and (b), respectively. The variables $v_1$, $v_2$, $v_3$, $v_4$ in the figure are \emph{character} variables. One special treatment we do in this paper is that we encoding characters as natural numbers. So $v_1$, $v_2$, $v_3$, $v_4$ are also integer variables. For example, from the formula $ y \in (12)^+$, we may derive $v_3=1$ and $v_4=2$.

@@
